---
editor_options: 
  markdown: 
    wrap: 72
---

# R

This R tutorial was adapted from
[Software Carpentry's Intro to R and RStudio Lessons](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html)

## Seeking Help with R

### Learning Objectives
-   To be able to read R help files for functions, data objects and
    special operators.
-   To be able to use CRAN task views to identify packages to solve a
    problem.

```{r, include=FALSE}
```

### Reading Help Files

R, and every package, provide help files for functions or data objects.
The general syntax to search for help on any function, "function_name",
from a specific function that is in a package loaded into your R session
is:

```{r, eval=FALSE}
?function_name
help(function_name)
```

For example take a look at the help file for `mean()`

```{r, eval=FALSE}
?mean()
```

This will load up a help page in RStudio.

Each help page is broken down into sections:

-   `Description`: A brief description of what a function does or what a
    data object contains.
-   `Usage`: For functions, shows how the function is called, including
    its arguments and their default values. For data objects, shows the
    name of the object and - sometimes how it is structured or loaded.
-   `Arguments`: For functions, describes each argument and the type of
    input it expects. This section is usually absent for data objects.
-   `Format`: Common for data objects; describes the structure of the
    object, such as the number of observations and variables, variable
    names, data types, and what each variable represents.
-   `Details`: Provides additional information about how the function
    works or important background information about the data object.
-   `Value`: For functions, describes what the function returns. This
    section is usually absent for data objects.
-   `Source`: Common for data objects; describes where the data came from
    or provides a reference to the original source.
-   `Examples`: Provides example code demonstrating how to use the
    function or work with the data object.

Different functions/objects might have different sections, but these are
the main ones you should be aware of.

### Tip: Running Examples

From within the function help page, you can highlight code in the
Examples and hit <kbd>Return</kbd> to run it in RStudio console. This
gives you a quick way to get a feel for how a function works.

### Tip: Reading Help Files

One of the most daunting aspects of R is the large number of functions
available. It would be prohibitive, if not impossible to remember the
correct usage for every function you use. Luckily, using the help files
means you don't have to remember that!

### Data Objects

R by default comes with a collection of built-in data objects from the
alphabets and months of the year to datasets collected about penguins
and tooth growth in guinea pigs. These objects provide convenient
examples for learning how to explore, manipulate, summarize, and
visualize different types of data in R. Additionally, many R packages
come with their own data objects, such as example datasets, reference
databases, lookup tables, and precomputed results. Understanding what
these data objects contain and how they are structured can be essential
for using certain packages effectively.

```{r, eval=FALSE}
?iris()
?LETTERS()
```

### Special Operators

To seek help on special operators, use quotes or backticks:

```{r, eval=FALSE}
?"<-"
?`<-`
```

### Getting Help with Packages

Many packages come with "vignettes": tutorials and extended example
documentation. Without any arguments, `vignette()` will list all
vignettes for all installed packages; `vignette(package="package-name")`
will list all available vignettes for `package-name`, and
`vignette("vignette-name")` will open the specified vignette.

If a package doesn't have any vignettes, you can usually find help by
typing `help("package-name")`.

RStudio also has a set of excellent
[cheatsheets](https://rstudio.com/resources/cheatsheets/) for many
packages.

### When You Remember Part of the Function Name

If you're not sure what package a function is in or how it's
specifically spelled, you can do a fuzzy search:

```{r, eval=FALSE}
??function_name
```

A fuzzy search is when you search for an approximate string match. For
example, you may remember that the function to set your working
directory includes "set" in its name. You can do a fuzzy search to help
you identify the function:

```{r, eval=FALSE}
??standarddeviation
```

### When You Have No Idea Where to Begin

If you don't know what function or package you need to use [CRAN Task
Views](https://cran.at.r-project.org/web/views) is a specially
maintained list of packages grouped into fields. This can be a good
starting point.

### Other Ways to Get Help

R's built-in documentation is not always enough, especially when you are
trying to solve a specific problem. Other useful sources of help
include:

-   Package websites: Popular packages often have dedicated websites
    containing documentation, tutorials, and examples.
-   Stack Overflow: Searching for the error message or problem you are
    encountering can often lead to solutions from other R users.
-   R community forums: Communities such as Posit Community provide a
    place to ask questions and discuss R-related problems.
-   GitHub: Package repositories can be useful for checking known
    issues, reporting bugs, and finding examples of how a package is
    being developed or used.
-   AI/LLMs: AI tools such as ChatGPT can help explain R concepts,
    interpret error messages, suggest functions or packages, and
    generate or troubleshoot code. However, AI-generated answers can be
    incorrect or outdated, so you should verify important information
    against official documentation and other reliable sources.

### Resources

-   [Quick R](https://www.statmethods.net/)
-   [RStudio cheat
    sheets](https://www.rstudio.com/resources/cheatsheets/)
-   [Cookbook for R](https://www.cookbook-r.com/)

## Data Structures, Data Types and Data frames

### Learning Objectives

-   To be able to identify the 5 main data types.
-   To begin exploring data frames, and understand how they are related
    to vectors and lists.
-   To be able to ask questions from R about the type, class, and
    structure of an object.
-   To understand the information of the attributes "names", "class",
    and "dim".

```{r, include=FALSE}
options(stringsAsFactors = FALSE)
```

One of R's most powerful features is its ability to work with **tabular
data**, such as data you might already have in a spreadsheet or CSV
file. Tabular data in R is commonly stored as a **data frame**, which
can be thought of as a collection of vectors arranged into columns. Each
vector represents a single column, while each row typically represents
an observation or record.

::: infobox
**What is a vector?**

In R, a vector is a sequence of values that all have the same data type.
Vectors are one of the fundamental data structures in R and form the
individual columns of a data frame.

Some of the most common data types you will encounter in R include:

-   `Double`: Numeric values that can contain decimals, such as 3.14 or
    10.5.
-   `Integer`: Whole numbers, such as 1, 5, or 100.
-   `Character`: Text values, such as "Hello" or "world".
-   `Logical`: Boolean values that can be either TRUE or FALSE.
:::

Let's start by creating a toy data frame containing information about
three different penguins.

```{r}
pg <- data.frame(
  species = c("Adelie", "Chinstrap", "Gentoo"),
  island = c("Torgersen", "Dream", "Biscoe"),
  bill_len = c(39.1, 46.5, 46.1),
  flipper_len = c(181, 192, 211),
  is_male = c(TRUE, FALSE, FALSE)
)
pg
```

We can begin exploring our dataset right away. Individual columns can be accessed by specifying the name of the data frame followed by the $ operator and the name of the column:

```{r, eval = F}
pg$bill_len
```


This returns the bill_len column as a vector. Because it is a numeric vector, we can perform mathematical operations directly on its values:

```{r, eval = F}
pg$bill_len * 2
```

Notice the operation is applied to every value in the vector.

### Data Types

To know the data types each of the column in our toy dataset, there are several methods to do this:

1. Use `str()` to examine the structure of the data frame

```{r}
str(pg)
```

`str()` provides a compact overview of the entire data frame, including its dimensions, column names, data types, and some example values. This is often one of the first functions you will use when exploring a new dataset.

2. Use `summary()` to summarize each column
```{r}
summary(pg)
```

`summary()` provides a summary appropriate for the type of data in each column. For numeric columns, this includes statistics such as the minimum, median, mean, and maximum. For other data types, the information displayed will differ.

3. Use `typeof()` to inspect the underlying type of a specific column

```{r, collapse = T}
typeof(pg$species)
typeof(pg$bill_len)
typeof(pg$flipper_len)
typeof(pg$is_male)
```

Notice that `flipper_len` is also stored as a double, even though all of its values are whole numbers. In R, entering a number such as 181 creates a double by default. To explicitly create an integer, the number is followed by L, for example:

```{r, collapse = T}
typeof(181)
typeof(181L)
```

For everyday data analysis, you will often use str() to quickly inspect an entire dataset and functions such as typeof() when you need to investigate a particular object in more detail.

### Manipulating data frames

Now that we know how to access parts of a data frame, let us try to manipulate the data stored within the data frame. There are many ways by which a data frame can be manipulated. For instance, we could append new columns, remove existing columns, add new observations or modify the values of an existing column.

1. Multiply the flipper length by two and store the vector as a new column
```{r}
flip_length_x2 <- pg$flipper_len * 2
pg$flip_length_x2 <- flip_length_x2
pg
```

2. Remove the newly added column from the data frame
```{r}
pg[,1:5]
```

3. Keep specific observations
```{r}
pg[c(1,3),1:5]
```

3. Modify an existing column
```{r}
pg$flipper_len <- pg$flipper_len * 2
pg
```
### Reading and Writing Data Frames

```{r}
write.table(
  pg, # the data object to be written; usually a matrix or data frame
  file = "penguins_dataset.csv", # file name
  quote = FALSE, # whether to quote any character columns
  sep = ",", # separator character e.g. , for CSV or \t for TSV
  row.names = FALSE, # whether to include row names in the output
  col.names = TRUE, # whether to include the column names in the output
)
```


```{r}
pg2 <- read.csv(
  "penguins_dataset.csv",
  header = TRUE
)
pg2
```

## Creating Plots with ggplot2

### Learning Objectives

- To be able to use ggplot2 to generate publication-quality graphics.
- To apply geometry, aesthetic, and statistics layers to a ggplot plot.
- To manipulate the aesthetics of a plot using different colors, shapes, and lines.
- To improve data visualization through transforming scales and paneling by group.
- To save a plot created with ggplot to disk.

Plotting our data is one of the best ways to
quickly explore it and the various relationships
between variables.

There are three main plotting systems in R,
the [base plotting system][base], the [lattice]
package, and the [ggplot2] package.

Today we'll be learning about the ggplot2 package, because
it is the most effective for creating publication-quality
graphics.

ggplot2 is built on the grammar of graphics, the idea that any plot can be
built from the same set of components: a **data set**,
**mapping aesthetics**, and graphical **layers**:

- **Data sets** are the data that you, the user, provide.

- **Mapping aesthetics** are what connect the data to the graphics.
  They tell ggplot2 how to use your data to affect how the graph looks,
  such as changing what is plotted on the X or Y axis, or the size or
  color of different data points.

- **Layers** are the actual graphical output from ggplot2. Layers
  determine what kinds of plot are shown (scatterplot, histogram, etc.),
  the coordinate system used (rectangular, polar, others), and other
  important aspects of the plot. The idea of layers of graphics may
  be familiar to you if you have used image editing programs
  like Photoshop, Illustrator, or Inkscape.

Let's start off building an example using the gapminder data from earlier.
The most basic function is `ggplot`, which lets R know that we're
creating a new plot. Any of the arguments we give the `ggplot`
function are the *global* options for the plot: they apply to all
layers on the plot.

```{r blank-ggplot, message=FALSE, fig.alt="Blank plot, before adding any mapping aesthetics to ggplot()."}
library("ggplot2")
penguins
```

Here we called `ggplot` and told it what data we want to show on
our figure. This is not enough information for `ggplot` to actually
draw anything. It only creates a blank slate for other elements
to be added to.

Now we're going to add in the **mapping aesthetics** using the
`aes` function. `aes` tells `ggplot` how variables in the **data**
map to *aesthetic* properties of the figure, such as which columns
of the data should be used for the **x** and **y** locations.

```{r ggplot-with-aes, message=FALSE, fig.alt="Plotting area with axes for a scatter plot of life expectancy vs GDP, with no data points visible."}
ggplot(data = penguins, mapping = aes(x = bill_len, y = flipper_len))
```

Here we told `ggplot` we want to plot the "gdpPercap" column of the
gapminder data frame on the x-axis, and the "lifeExp" column on the
y-axis. Notice that we didn't need to explicitly pass `aes` these
columns (e.g. `x = gapminder[, "gdpPercap"]`), this is because
`ggplot` is smart enough to know to look in the **data** for that column!

The final part of making our plot is to tell `ggplot` how we want to
visually represent the data. We do this by adding a new **layer**
to the plot using one of the **geom** functions.

```{r lifeExp-vs-gdpPercap-scatter, warning = F, message=FALSE, fig.alt="Scatter plot of life expectancy vs GDP per capita, now showing the data points."}
ggplot(data = penguins, mapping = aes(x = bill_len, y = flipper_len)) +
  geom_point()
```

Here we used `geom_point`, which tells `ggplot` we want to visually
represent the relationship between **x** and **y** as a scatterplot of points.

```{r lifeExp-vs-gdpPercap-scatter, warning = F, message=FALSE, fig.alt="Scatter plot of life expectancy vs GDP per capita, now showing the data points."}
ggplot(data = penguins, mapping = aes(x = body_mass, y = flipper_len, color = species)) +
  geom_point()
```

```{r lifeExp-vs-gdpPercap-scatter, warning = F, message=FALSE, fig.alt="Scatter plot of life expectancy vs GDP per capita, now showing the data points."}
ggplot(data = penguins, mapping = aes(x = flipper_len, fill = species)) +
  geom_histogram()
```

